Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-05-16 paper/paper.qmd
Adapt methods section from https://github.com/SchlossLab/Tomkovich_CDI_clinical_XXXX_2021

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
bibliography: references.bib
fontsize: 11pt
date: last-modified
date-format: medium
format:
  manuscript-pdf:
    mainfont: Helvetica
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting Severe _C. difficile_ infections
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
abstract: TODO
importance: TODO
keywords: _C. difficile_ infection, supervised machine learning
---

@@

(@fig-features)

@@

## Sample collection and 16S rRNA gene amplicon sequencing

[@schloss_introducing_2009]

@@

[@topcuoglu_mikropml_2021]

@@

## Code and data availability

[@sovacool_schtools_2022]

>>>>

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
bibliography: references.bib
fontsize: 11pt
date: last-modified
date-format: medium
format:
  manuscript-pdf:
    mainfont: Helvetica
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
abstract: TODO
importance: TODO
keywords: _C. difficile_ infection, supervised machine learning
---

@@

Most important OTUs contributing to model performance (@fig-features)

@@

## Sample collection

This study was approved by the University of Michigan Institutional Review
Board. All patient samples were collected by the University of Michigan Health
System from January 2016 through December 2017.
Stool samples that had unformed stool consistency were tested for *C. difficile* by
the clinical microbiology lab with a two-step algorithm that included detection
of *C. difficile* glutamate dehydrogenase and toxins A and B by enzyme
immunoassay with reflex to PCR for the *tcdB* gene when results were discordant.
1,517 stool samples were collected from patients diagnosed with a CDI.
Leftover stool samples that were sent to the clinical microbiology lab were
collected and split into different aliquots.
For 16S sequencing, the aliquot of stool was resuspended in
DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.
Only the first CDI sample per patient was used for subsequent ML analyses such
that no patient is represented more than once, resulting in a dataset of 1,191
samples.

## 16S rRNA gene amplicon sequencing

Samples stored in DNA genotek buffer were thawed from the -80&deg;C, vortexed,
and then transferred to a 96-well bead beating plate for DNA extractions.
DNA was extracted using the DNeasy
Powersoil HTP 96 kit (Qiagen) and an EpMotion 5075 automated pipetting system
(Eppendorf).
The V4 region of the 16S rRNA gene was amplified with the AccuPrime
Pfx DNA polymerase (Thermo Fisher Scientific) using custom barcoded primers, as
previously described [@kozich_development_2013].
Each library preparation plate for
sequencing contained a negative control (water) and mock community control
(ZymoBIOMICS microbial community DNA standards).
The PCR amplicons were
normalized (SequalPrep normalizatin plate kit from Thermo Fisher Scientific),
pooled and quantified (KAPA library quantification kit from KAPA Biosystems),
and sequenced with the MiSeq system (Illumina).

All sequences were processed with mothur (v1.43) using the MiSeq SOP protocol
[@schloss_introducing_2009; @kozich_development_2013].
Paired sequencing reads were combined and
aligned with the SILVA (v132) reference database [@quast_silva_2013] and taxonomy was
assigned with a modified version of the Ribosomal Database Project reference
sequences (v16) [@cole_ribosomal_2014].
Samples were rarefied to 5,000 sequences per sample, repeated 1,000
times for alpha and beta diversity analysis.

@@

mikropml R package [@topcuoglu_mikropml_2021]

@@

## Code availability

The complete workflow, code, and supporting files required to reproduce this
manuscript with accompanying figures is available at
https://github.com/SchlossLab/severe-CDI.
<!-- TODO update GitHub URL once accepted to journal -->

The steps for machine learning, making figures, and rendering the manuscript
were defined with Snakemake [@koster_snakemake_2012] using a custom version of the
mikropml Snakemake workflow [@sovacool_mikropml_2023].
Dependencies were managed with conda environments.
In addition to the software already cited above, other packages used in the
creation of this manuscript include
schtools [@sovacool_schtools_2022],
Quarto,
ggtext [@wilke_ggtext_2020],
ggsankey [@sjoberg_ggsankey_2022],
cowplot [@wilke_cowplot_2020], and
vegan [@oksanen_vegan_2022].

## Data availability

The 16S rRNA sequencing data have been deposited in the National Center for
Biotechnology Information Sequence Read Archive
(BioProject Accession no. PRJNA729511).
